Summary of tumor cell classification with `SingleR` for `r params$library_id`

### 2024-07-22 analyses/cell-type-ewings/template_notebooks/singler-workflow/01-singler-results.Rmd
add in singler vs. aucell comparison

```{r}
# original library annotation 
orig_tumor <- glue::glue("tumor-{params$library_id}")

# generate classification df to use for plots 
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2) |> 
  # join with previous annotations, singler results, and gene set scores 
  dplyr::left_join(singler_results_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes") |> 
  dplyr::mutate(
    # first remove other library IDs from tumor annotations
    # we only care about whether or not it was from the original library or a new library
    singler_tumor_annotation = dplyr::case_when(
      singler_tumor_annotation == orig_tumor ~ "tumor_original",
      stringr::str_detect(singler_tumor_annotation, "tumor") ~ "tumor_new",
      .default = singler_tumor_annotation
    ),
    # get the top cell types for plotting later 
    singler_tumor_annotation_lumped = dplyr::if_else(is.na(singler_tumor_annotation), "Not labeled", singler_tumor_annotation) |>
          # there are warnings because we have some that have less than 7 cell types (ones that only did tumor/normal)
          # this is why we ignore warnings in this chunk 
          forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
          forcats::fct_infreq() |>
          forcats::fct_relevel("All remaining cell types", after = Inf), 
    # make a label that combines tumor_new with tumor_original
    singler_combined_tumor = dplyr::if_else(
      singler_tumor_annotation_lumped %in% c("tumor_original", "tumor_new"), 
      "tumor", 
      singler_tumor_annotation_lumped
      ) |> 
      forcats::fct_relevel("tumor", after = 0) |> 
      forcats::fct_relevel("All remaining cell types", after = Inf),
    # make a label that is just tumor or normal
    tumor_normal_classification = dplyr::if_else(singler_combined_tumor == "tumor", "tumor", "normal") |> 
      forcats::fct_relevel("tumor", after = 0)
  )
```

@@

# join sum expression columns with classification df 
classification_df <- classification_df |> 
  dplyr::left_join(gene_exp_df)
```

@@

## UMAP of all cell type annotations

@@

```{r, fig.height=8, fig.width=8}
ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = singler_tumor_annotation_lumped)) +
  # set points for all "other" points
  geom_point(
    data = dplyr::select(
      classification_df, -singler_tumor_annotation_lumped
    ),
    color = "gray80",
    alpha = 0.5,
    size = 0.1
  ) +
  # set points for desired cell type
  geom_point(size = 0.1, alpha = 0.5) +
  facet_wrap(
    vars(singler_tumor_annotation_lumped),
    ncol = 3
  ) +
  scale_color_brewer(palette = "Dark2") +
  # remove axis numbers and background grid
  scale_x_continuous(labels = NULL, breaks = NULL) +
  scale_y_continuous(labels = NULL, breaks = NULL) +
  guides(
    color = guide_legend(
      title = "Cell type",
      # more visible points in legend
      override.aes = list(
        alpha = 1,
        size = 1.5
      )
    )) +
  theme(
        legend.position = c(0.67, 0.33),
        legend.justification = c("left", "top"),
        legend.title.align = 0.5,
        # use slightly smaller legend text, which helps legend fit and prevents
        #  long wrapped labels from bunching up
        legend.text = element_text(size = rel(0.85)),
        legend.key.height = unit(0.75, "cm"),
        aspect.ratio = 1
      )
```

## Evaluate output from `SingleR`

The next three sections will evaluate the annotations by looking at expression of marker genes and gene set scores.

@@

In this section we show just the cells that are considered tumor (including both `tumor_original` and `tumor_new`) and normal, lumping all non-tumor cell types together.

@@

```{r}
full_celltype_heatmap(classification_df, marker_gene_columns, "tumor_normal_classification")
```

```{r}
plot_density(classification_df,
             "tumor_sum",
             "tumor_normal_classification")
```

@@

```{r}
full_celltype_heatmap(classification_df, geneset_columns, "tumor_normal_classification")
```

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "tumor_normal_classification")
  })
```

@@

In this section we show all tumor cells together (including both `tumor_original` and `tumor_new`) and the top 5 most represented normal cell types.

@@

```{r}
full_celltype_heatmap(classification_df, marker_gene_columns, "singler_combined_tumor")
```

```{r}
marker_gene_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_combined_tumor")
  })
```

@@

```{r}
full_celltype_heatmap(classification_df, geneset_columns, "singler_combined_tumor")
```

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_combined_tumor")
  })
```

### Compare newly annotated tumor cells to existing tumor cells

In this section we show both tumor cell types, `tumor_original` and `tumor_new`, separately and the top 5 most represented normal cell types.

@@

```{r}
full_celltype_heatmap(classification_df, marker_gene_columns, "singler_tumor_annotation_lumped")
```

```{r}
marker_gene_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_tumor_annotation_lumped")
  })
```

@@

```{r}
full_celltype_heatmap(classification_df, geneset_columns, "singler_tumor_annotation_lumped")
```

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_tumor_annotation_lumped")
  })
```

>>>>

```{r}
# generate classification df to use for plots 
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1 and get rid of excess columns
  dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
  # join with previous annotations, singler results, and gene set scores 
  dplyr::left_join(singler_results_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes") |> 
  dplyr::mutate(
    # set factor order for aucell 
    aucell_annotation = forcats::fct_relevel(aucell_annotation, "tumor", after= 0), 
    # first remove other library IDs from tumor annotations
    # we only care about whether or not it was from the original library or a new library
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |> 
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later 
    singler_lumped = singler_annotation |>
          # there are warnings because we have some that have less than 7 cell types (ones that only did tumor/normal)
          # this is why we ignore warnings in this chunk 
          forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
          forcats::fct_infreq() |>
          forcats::fct_relevel("All remaining cell types", after = Inf), 
    # make a label that is just tumor or normal
    singler_tumor_normal = dplyr::if_else(!(singler_annotation %in% c("tumor", "unknown")), "normal", singler_annotation) |> 
      forcats::fct_relevel("tumor", after = 0), 
    # get the consensus between aucell and singler 
    consensus = dplyr::case_when(
      singler_annotation == "tumor" & aucell_annotation == "tumor" ~ "tumor_both",
      singler_annotation == "tumor" & aucell_annotation != "tumor" ~ "tumor_singler",
      singler_annotation != "tumor" & aucell_annotation == "tumor" ~ "tumor_aucell",
      .default = "normal"
    ) |> 
      forcats::fct_relevel("normal", after = Inf)
  )
```

@@

# join sum expression columns with classification df 
classification_df <- classification_df |> 
  dplyr::left_join(gene_exp_df, by = "barcodes")
```

@@

## UMAPs of all cell type annotations

@@

```{r, fig.height=8, fig.width=8}
plot_faceted_umap(classification_df, singler_lumped)
```

## Comparison between `SingleR` and `AUCell`

Below we show the cells that are annotated as tumor by both methods, just `AUCell` and just `SingleR` along with all normal cells.

```{r}
plot_faceted_umap(classification_df, consensus)
```

We then calculate the confusion matrix between tumor and normal cells called in `SingleR` and `AUCell`. 
The rows here correspond to the `AUCell` annotation and columns are from `SingleR`.

```{r}
caret_df <- classification_df |> 
  dplyr::filter(singler_tumor_normal != "unknown") |> 
  dplyr::mutate(singler_tumor_normal = droplevels(singler_tumor_normal, "unknown"))

caret::confusionMatrix(
  table(
    caret_df$aucell_annotation,
    caret_df$singler_tumor_normal
  )
)
```

## Validate output from `SingleR`

The next three sections will evaluate the annotations output obtained from using `SingleR` by looking at expression of marker genes and gene set scores.

@@

In this section we show just the cells that are considered tumor and normal, lumping all non-tumor cell types together.

@@

```{r}
full_celltype_heatmap(classification_df, marker_gene_columns, "singler_tumor_normal")
```

```{r}
plot_density(classification_df,
             "tumor_sum",
             "singler_tumor_normal")
```

@@

```{r}
full_celltype_heatmap(classification_df, geneset_columns, "singler_tumor_normal")
```

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_tumor_normal")
  })
```

@@

In this section we show all tumor cells and the top 5 most represented normal cell types.

@@

```{r}
full_celltype_heatmap(classification_df, marker_gene_columns, "singler_lumped")
```

```{r}
marker_gene_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_lumped")
  })
```

@@

```{r}
full_celltype_heatmap(classification_df, geneset_columns, "singler_lumped")
```

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_lumped")
  })
```

### Tumor cells annotatd by `SingleR` and `AUCell`

Here we compare the marker gene expression and gene set scores for cells annotated as tumor by both methods, just `SingleR`, or just `AUCell` and compare to all normal cells.

@@

```{r}
full_celltype_heatmap(classification_df, marker_gene_columns, "consensus")
```

```{r}
plot_density(classification_df,
             "tumor_sum",
             "consensus")
```

@@

```{r}
full_celltype_heatmap(classification_df, geneset_columns, "consensus")
```

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "consensus")
  })
```
